\begin{thebibliography}{99}

\bibitem[Hart(1975)]{hart1975}
Hart, M.~H. (1975). An explanation for the absence of extraterrestrials on Earth. \textit{Q. J. R. Astron. Soc.}, 16, 128--135.

\bibitem[Hanson(1998)]{hanson1998}
Hanson, R. (1998). The Great Filter---Are we almost past it? \textit{Working paper}, George Mason University.

\bibitem[Bostrom(2002)]{bostrom2002}
Bostrom, N. (2002). Existential risks: Analyzing human extinction scenarios. \textit{J. Evol. Tech.}, 9(1), 1--31.

\bibitem[Bostrom(2014)]{bostrom2014}
Bostrom, N. (2014). \textit{Superintelligence}. Oxford University Press.

\bibitem[Tainter(1988)]{tainter1988}
Tainter, J.~A. (1988). \textit{The Collapse of Complex Societies}. Cambridge University Press.

\bibitem[Turchin(2003)]{turchin2003}
Turchin, P. (2003). \textit{Historical Dynamics}. Princeton University Press.

\bibitem[Turchin(2016)]{turchin2016}
Turchin, P. (2016). \textit{Ages of Discord}. Beresta Books.

\bibitem[Scheidel(2017)]{scheidel2017}
Scheidel, W. (2017). \textit{The Great Leveler}. Princeton University Press.

\bibitem[Smil(2017)]{smil2017}
Smil, V. (2017). \textit{Energy and Civilization: A History}. MIT Press.

\bibitem[Piketty(2014)]{piketty2014}
Piketty, T. (2014). \textit{Capital in the Twenty-First Century}. Harvard University Press.

\bibitem[Acemoglu and Robinson(2012)]{acemoglu2012}
Acemoglu, D. \& Robinson, J.~A. (2012). \textit{Why Nations Fail}. Crown Business.

\bibitem[Putnam(2000)]{putnam2000}
Putnam, R.~D. (2000). \textit{Bowling Alone}. Simon \& Schuster.

\bibitem[Chetty et~al.(2017)]{chetty2017}
Chetty, R., Grusky, D., Hell, M., Hendren, N., Manduca, R., \& Narang, J. (2017). The fading American dream. \textit{Science}, 356(6336), 398--406.

\bibitem[Storn and Price(1997)]{storn1997}
Storn, R. \& Price, K. (1997). Differential evolution. \textit{J. Global Optim.}, 11(4), 341--359.

\bibitem[Olson(1965)]{olson1965}
Olson, M. (1965). \textit{The Logic of Collective Action}. Harvard University Press.

\bibitem[De~Soto(1989)]{de_soto1989}
De Soto, H. (1989). \textit{The Other Path}. Harper \& Row.

\bibitem[Webb(2002)]{webb2002}
Webb, S. (2002). \textit{If the Universe Is Teeming with Aliens... Where Is Everybody?} Copernicus.

\bibitem[Eichengreen(2011)]{eichengreen2011}
Eichengreen, B. (2011). \textit{Exorbitant Privilege}. Oxford University Press.

\bibitem[Perez(2002)]{perez2002}
Perez, C. (2002). \textit{Technological Revolutions and Financial Capital}. Edward Elgar.

\bibitem[Gaddy and Ickes(2015)]{gaddy2015}
Gaddy, C.~G. \& Ickes, B.~W. (2015). \textit{Bear Traps on Russia's Road to Modernization}. Routledge.

\bibitem[Varoufakis(2023)]{varoufakis2023}
Varoufakis, Y. (2023). \textit{Technofeudalism: What Killed Capitalism}. Penguin.

\bibitem[Ward-Perkins(2005)]{wardperkins2005}
Ward-Perkins, B. (2005). \textit{The Fall of Rome and the End of Civilization}. Oxford University Press.

\bibitem[Wickham(2009)]{wickham2009}
Wickham, C. (2009). \textit{The Inheritance of Rome: A History of Europe from 400 to 1000}. Viking.

\bibitem[McCormick(2001)]{mccormick2001}
McCormick, M. (2001). \textit{Origins of the European Economy: Communications and Commerce, AD 300--900}. Cambridge University Press.

\end{thebibliography}
